Numerical simulation and entropy dissipative cure of the carbuncle instability for the shallow water circular hydraulic jump

### 2021-01-14 numerics_paper/version2/numerical_chj.tex
Minor edits to sections 4-5.

\subsection{Wave propagation methods}\label{sec:waveprop}
The numerical solution $q_i$ represents the average value of the set of
conserved quantities over cell $i$, which extends from $x_\imh$ to $x_\iph$.
At each time step and at each interface $x_\imh$, we approximately solve the
Riemann problem with initial states $(q^n_{i-1},q^n_i)$.  The approximate
solution consists of a set of traveling discontinuities or waves $\W^p_\imh$
and corresponding speeds $s^p_\imh$.
We use the wave propagation framework developed by LeVeque \cite{leveque1997wave, leveque2002finite}
and employed in the Clawpack software \cite{clawpack,pyclaw-sisc}, which implements the LWL scheme
\begin{align}\label{second-order_via_fluct}
  q_i^{n+1} = q_i^n-\frac{\Delta t}{\Delta x}\left[\apdq_\imh+\amdq_\iph\right]
  -\frac{\Delta t}{\Delta x}\left[\tilde{F}_{i+1/2}-\tilde{F}_{i-1/2}\right],
\end{align}
Upon defining $z^+:=\frac{1}{2}(z+|z|)$ and $z^-:=\frac{1}{2}(z-|z|)$,
the fluctuations are given by
\begin{align}\label{fluct}
  \apdq_\imh := \sum_p\left(s_{i-1/2}^p\right)^+\W_{i-1/2}^p, \qquad
  \amdq_\iph := \sum_p\left(s_{i+1/2}^p\right)^-\W_{i+1/2}^p, \\
\end{align}
and represent the effect (to first
order accuracy) of waves coming from the solution of the Riemann problem at
each of the neighboring interfaces.  Meanwhile, $\Ft_{i\pm 1/2}$ are correction
fluxes that make the method second-order accurate:
\begin{align*}
  \tilde{F}_{i\pm 1/2}=\frac{1}{2}\sum_p|s_{i\pm 1/2}^p|\left(1-\frac{\Delta t}{\Delta x}|s_{i\pm 1/2}^p|\right)\tilde\W_{i\pm 1/2}^p.
\end{align*}
Here $\tilde{\W}_{i\pm 1/2}^p$ is a TVD-limited version of $\W_{i\pm 1/2}^p$.
We will sometimes work with
the first-order method obtained by setting the correction fluxes to zero:
\begin{align}\label{first-order_via_fluct}
  q_i^{n+1} = q_i^n-\frac{\Delta t}{\Delta x}\left[\apdq_\imh+\amdq_\iph\right]
\end{align}
which, for the conservative Riemann solvers we use herein, can also be written
as a first-order flux-differencing scheme:
\begin{align}\label{first-order_FV}
  q_i^{n+1}=q_i^n-\frac{\Delta t}{\Delta x}\left[\bfF_{i+1/2}-\bfF_{i-1/2}\right],
\end{align}
where $\bfF_{i+1/2}$ is a numerical approximation of the flux at $x_\iph$.

@@

%We close this section by remarking that the fluctuations associated with Rusanov's solver
%can be expressed in terms of waves and speeds from Roe's solver. To see this, first consider \eqref{rus_cons}
%and rewrite the right-going fluctuation as follows:
%\begin{align}\label{rus_rs_aux}
%  \A^{+,\Rus}\Delta Q_{i-1/2}:=\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\max}
%  = \bff(q_{i})-\bff(q_{i-1}) + \lambda_{i-1/2}^{\max}\left(\W_{i-1/2}^{1,\Rus}+\W_{i-1/2}^{2,\Rus}\right)
%  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus}.
%  %\A^{+,\Rus}\Delta Q_{i-1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p
%\end{align}
%From \eqref{system_for_alphas} and \eqref{rusanov_waves} we get
%$\W^{1,\Rus}_{i-1/2}+\W^{2,\Rus}_{i-1/2}=q_i-q_{i-1}=\sum_p\W_{i-1/2}^p$.
%Using \eqref{rs_conservation}, \eqref{rus_rs_aux} becomes
%\begin{align*}
%  \lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\max}
%  = \sum_p \left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p
%  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus},
%\end{align*}
%which implies that
%\begin{subequations}\label{fluct_rus}
%\begin{align}
%  \A^{+,\Rus}\Delta Q_{i-1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p,
%\end{align}
%and similarly,
%\begin{align}
%  \A^{-,\Rus}\Delta Q_{i+1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i+1/2}^p - \lambda_{i+1/2}^{\max}\right)\W_{i+1/2}^p.
%\end{align}
%\end{subequations}

@@

\subsection{The entropy residual}
Let $\entvar(\bfq)$ and $\entflux(\bfq)$ denote the entropy variable and entropy flux.
Based on \cite{guermond2011entropy}, we consider
\begin{align}\label{ent_residual}
  \int_{S_i} \entvar(\bfq) \cdot \left[ \frac{\partial \bfq}{\partial t} + \nabla\cdot \bff(\bfq)\right]d\bfx
  =\int_{S_i} \left[\frac{\partial \eta(\bfq)}{\partial t} + \entvar(\bfq) \cdot \nabla\cdot \bff(\bfq)\right] d\bfx
\end{align}
as a measurement of the entropy production in cell $i$ (here $S_i=[x_\imh,x_\iph]$.
To avoid the need to compute the time derivative of the entropy, we follow
\cite{guermond2018second, guermond2018well} and use
$\int \frac{\partial\eta(\bfq)}{\partial t}d\bfx=-\int \nabla\cdot\entflux(\bfq) d\bfx$,
which holds in smooth regions. Then we define
\begin{align}\label{Ri}
  \theta_i := \frac{R_i}{D_i},
\end{align}
with
\begin{align*}
  R_i=
  \left|\int_{S_i} \left[\entvar(\bfq) \cdot \nabla\cdot \bff(\bfq) - \nabla\cdot\entflux(\bfq) \right] d\bfx\right|,
\end{align*}
and $D_i$ being an upper bound of $R_i$ that acts as normalization so that $0\leq \theta_i\leq 1$.
Note that $R_i\approx 0$ if $\bfq$ is smooth in $S_i$.
In our implementation, we use
\begin{align*}
  R_i\approx \left|\entvar(\bfq_i)\cdot \int_{S_i}\nabla\cdot \bff(\bfq)d\bfx
  -\int_{S_i}\nabla\cdot\entflux(\bfq) d\bfx\right|,
  \quad
  %D_i = ||\entvar(\bfq_i)||_{\ell^2}\left|\left|\int_{S_i}\nabla\cdot\bff(\bfq)d\bfx\right|\right|_{\ell^2}
  %+\left|\int_{S_i}\nabla\cdot\bfq(\bfq)d\bfx\right|.
  D_i = \sum_{k=1}^{d+1}|\entvar_k(\bfq_i)|\left|\int_{S_i}\left(\nabla\cdot\bff(\bfq)\right)_kd\bfx\right|
  +\left|\int_{S_i}\nabla\cdot\entflux(\bfq)d\bfx\right|,
\end{align*}
where $({\bf z})_k$ denotes the $k$-th component of ${\bf z}\in\mathbb{R}^{d+1}$ and $d$ is the number of physical
dimensions. In the next section, we use $\theta_i$ to blend Rusanov's and Roe's Riemann solvers.

@@

To implement a second-order LWL scheme based on this solver, in addition to
these fluctuations we must define a set of waves and corresponding speeds.
Using only the waves from the Roe solver, it is in general not possible
to choose speeds that yield the fluctuations \eqref{ev_fluctuations}.
%In the framework of LeVeque, for a given interface $i+1/2$,
%we would need to consider a wave speed $s_{i+1/2}^p$ so that
%\begin{align*}
%  \hat \lambda_{i+1/2}^p+\lambda_{i+1/2}^p = s_{i+1/2}^p+|s_{i+1/2}^p|, \qquad
%  \hat \lambda_{i+1/2}^p-\lambda_{i+1/2}^p = s_{i+1/2}^p-|s_{i+1/2}^p|,
%\end{align*}
%which is not possible to obtain in general (since the problem is overdetermined).
Instead, we consider
$s_{i+1/2}^p=\sgn(\hat\lambda_{i+1/2}^p)\lambda_{i+1/2}^p$
 and study the consequences of this choice once the second-order corrections are applied.
Here $\sgn(z)$ is the sign function of $z$.
The second-order LWL method for the one-dimensional problem is given by
\begin{align}\label{second-order}
  q_i^{n+1}=q_i^n
  -\frac{\Delta t}{\Delta x}\left[\A^-\Delta Q_{i+1/2}+\A^+\Delta Q_{i-1/2}\right]
  -\frac{\Delta t}{\Delta x}\left[\tilde F_{i+1/2}-\tilde F_{i-1/2}\right],
\end{align}
where
\begin{align*}
  \tilde F_{i+1/2} = \frac{1}{2}\sum_p\lambda_{i+1/2}^p
  \left(1-\frac{\Delta t}{\Delta x}\lambda_{i+1/2}^p\right)\tilde\W_{i+1/2}^p
\end{align*}
are the flux corrections with $\tilde\W_{i+1/2}^p$ being a limited version of $\W_{i+1/2}^p$.

To understand the effect of choosing $s_{i+1/2}^p=\sgn(\hat\lambda_{i+1/2}^p)\lambda_{i+1/2}^p$
as the speed of $\W_{i+1/2}^p$,
let us assume that no limiters are applied; i.e., $\tilde\W_{i\pm 1/2}^p\equiv \W_{i\pm 1/2}^p$.
By doing this, equation \eqref{second-order} can be written as
\begin{align*}
  q_i^{n+1}=q_i^n
  &-\frac{\Delta t}{2 \Delta x}
  \sum_p
  \left[\left(\hat\lambda_{i+1/2}^p -\frac{\Delta t}{\Delta x}(\lambda_{i+1/2}^p)^2 \right)\W_{i+1/2}^p
  +
  \left(\hat\lambda_{i-1/2}^p +\frac{\Delta t}{\Delta x}(\lambda_{i-1/2}^p)^2 \right)\W_{i-1/2}^p\right],
\end{align*}
which corresponds to the standard Lax-Wendroff method written in terms of fluctuation-like quantities.
Note that if $\theta_{i\pm 1/2}=0$, we recover the standard second-order Lax-Wendroff method based on
Roe's Riemann solver.
In the numerical experiments in \S\ref{sec:num}, we obtain second-order experimental
order of convergence (EOC) for problems with smooth solution, which is expected since
$\theta_{i\pm 1/2}\approx 0$ in smooth regions.
Similarly, if $\theta_{i\pm 1/2}=1$, we recover the Lax-Wendroff method based on Rusanov's Riemann solver,
which can be seen by writing the fluctuations in Rusanov's method in terms of Roe's waves and speeds.

To see this, first consider \eqref{rus_cons} and rewrite the right-going fluctuation as follows:
\begin{align}\label{rus_rs_aux}
  \A^{+,\Rus}\Delta Q_{i-1/2}:=\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\max}
  = \bff(\bfq_{i})-\bff(\bfq_{i-1}) + \lambda_{i-1/2}^{\max}\left(\W_{i-1/2}^{1,\Rus}+\W_{i-1/2}^{2,\Rus}\right)
  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus}.
  %\A^{+,\Rus}\Delta Q_{i-1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p
\end{align}
From \eqref{system_for_alphas} and \eqref{rusanov_waves} we get
$\W^{1,\Rus}_{i-1/2}+\W^{2,\Rus}_{i-1/2}=\bfq_i-\bfq_{i-1}=\sum_p\W_{i-1/2}^p$.
Using \eqref{rs_conservation}, \eqref{rus_rs_aux} becomes
\begin{align*}
  \lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\max}
  = \sum_p \left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p
  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus},
\end{align*}
which implies that
\begin{subequations}\label{fluct_rus}
\begin{align}
  \A^{+,\Rus}\Delta Q_{i-1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p,
\end{align}
and similarly,
\begin{align}
  \A^{-,\Rus}\Delta Q_{i+1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i+1/2}^p - \lambda_{i+1/2}^{\max}\right)\W_{i+1/2}^p.
\end{align}
\end{subequations}

@@

We consider a total of six problems. We start with a one-dimensional Riemann problem over a dry bed.
For this problem we use the first-order method \eqref{first-order_via_fluct},
to avoid negative depth values. Although only Rusanov's solver is proven to guarantee
positivity, we do not get negative values for the water height with any of the first-order methods.
In \S \ref{sec:steady_outflow} we consider a one-dimensional problem with a smooth steady state solution. We observe the expected
second-order accuracy of method \eqref{second-order_via_fluct} with Roe's and the blended solvers. Using
Rusanov's Riemann solver degrades the accuracy to first-order. The overall accuracy of the blended solver
is not degraded since the extra dissipation is not applied in smooth regions.
In \S\ref{sec:rp_wet_bed} and \S\ref{sec:1D_chj} we consider one-dimensional problems with strong shocks. We observe similar accuracy with the
blended solver than with Roe's solver; that is, inducing entropy dissipation near the shocks do not degrade the accuracy
of the underlying Roe's solver. This extra dissipation, however, prevents the formation of carbuncles in other experiments,
which we demonstrate in \S \ref{sec:bow_shock} and \S\ref{sec:2D_chj}.
The main focus of this work is in the formation of carbuncle instabilities in the two-dimensional CHJ. We present
an extensive set of experiments about this problem in \S \ref{sec:2D_chj}.
We consider not only Roe's, Rusanov's and the blended solvers, but also the HLLE solver and a solver proposed in \cite{kemm2014note},
which is designed to remove the carbuncle instabilities in the shallow water equations.

\subsection{Dam break problem on a dry bed}\label{sec:rp_dry_bed}
We start with the one-dimensional dam break problem on a dry bed.
This problem is a canonical example that demonstrates the `entropy glitch' of Roe's solver
at transonic rarefactions. Since the baseline Roe's solver we use (see \S\ref{sec:roe}) does not contain
an entropy fix, it is important to demonstrate that the blended Riemann solver fixes the entropy glitch.
We follow the setup in \cite[\S 4.1.2]{delestre2013swashes}.
The domain is given by $\Omega=(0,L=10)$, and the initial condition is $hu(x,0)=0$ and
\begin{align*}
  h(x,0)=
  \begin{cases}
    h_l, & \mbox{ if }, x\leq x_0=5 \\
    h_r, &\mbox{ otherwise},
  \end{cases}
\end{align*}
with $h_l=5\times 10^{-3}$ and $h_r=0$. In our experiments we use $h_r=1\times 10^{-15}$ to avoid division by zero.
The exact solution, which can also be found in \cite{delestre2013swashes} and references therein, is
\begin{align*}
  h(x,t) =
  \begin{cases}
    h_l, \\
    \frac{4}{9g}\left(\sqrt{gh_l}-\frac{x-x_0}{2t}\right)^2, \\
    0,
  \end{cases}
\quad
  u(x,t) =
  \begin{cases}
    0, &\mbox{ if } x\leq x_A(t), \\
    \frac{2}{3}\left(\sqrt{gh_l}+\frac{x-x_0}{t}\right), & \mbox{ if } x_A(t) < x\leq x_B(t), \\
    0, &\mbox{ if } x_B(t)\leq x,
  \end{cases}
\end{align*}
with $x_A(t)=x_0-t\sqrt{gh_l}$ and $x_B(t)=x_0+2t\sqrt{gh_l}$.
We use $g=1$ and solve the problem up to $t=10$.
In figures \ref{fig:rp_dry_bed_roe}-\ref{fig:rp_dry_bed_blended},
we show the solution using method \eqref{first-order_via_fluct}
with Roe's solver, Rusanov's solver and the blended solver, respectively.
The entropy glitch, which is manifested as a non-physical shock at $x=x_0$,
is present with Roe's solver. Using Rusanov's and the blended solvers
fixes the entropy glitch.
In Table \ref{table:rp_dry_bed}, we summarize the results of a convergence test.
Note that using the blended solver leads to more accurate results.

@@

For this particular problem, using method \eqref{first-order_via_fluct} with the blended Riemann solver
leads to $\lambda_{i\pm 1/2}^{\min}=0$ for all the experiments, but this is not true in general.
Finally, let us demonstrate the effect of $\lambda_{i\pm 1/2}^{\min}$ by imposing $\theta_{i}=0$ in \eqref{lambda_p},
which is equivalent to using Roe's solver with extra dissipation given by $\lambda_{i\pm 1/2}^{\min}$.
In Figure \ref{fig:rp_dry_bed_roe_with_lmin}, we show the solution using different refinements,
and in the last column of Table \ref{table:rp_dry_bed}, we summarize the results of a convergence test.
Note that the entropy glitch is highly reduced.
To remove completely the entropy glitch we could add high-order entropy dissipation following \cite{tadmor2003entropy}
and references therein.

>>>>

\subsection{Wave propagation methods}\label{sec:waveprop}
The numerical solution $q_i$ represents the average value of the set of
conserved quantities over cell $i$, which extends from $x_\imh$ to $x_\iph$.
At each time step and at each interface $x_\imh$, we approximately solve the
Riemann problem with initial states $(q^n_{i-1},q^n_i)$.  The approximate
solution consists of a set of traveling discontinuities or waves $\W^p_\imh$
and corresponding speeds $s^p_\imh$.
We use the wave propagation framework developed by LeVeque \cite{leveque1997wave, leveque2002finite}
and employed in the Clawpack software \cite{clawpack,pyclaw-sisc}, which implements the LWL scheme
\begin{align}\label{second-order_via_fluct}
  q_i^{n+1} = q_i^n-\frac{\Delta t}{\Delta x}\left[\apdq_\imh+\amdq_\iph\right]
  -\frac{\Delta t}{\Delta x}\left[\tilde{F}_{i+1/2}-\tilde{F}_{i-1/2}\right],
\end{align}
Upon defining $z^+:=\frac{1}{2}(z+|z|)$ and $z^-:=\frac{1}{2}(z-|z|)$,
the fluctuations are given by
\begin{align}\label{fluct}
  \apdq_\imh := \sum_p\left(s_{i-1/2}^p\right)^+\W_{i-1/2}^p, \qquad
  \amdq_\iph := \sum_p\left(s_{i+1/2}^p\right)^-\W_{i+1/2}^p, \\
\end{align}
and represent the effect (to first
order accuracy) of waves coming from the solution of the Riemann problem at
each of the neighboring interfaces.  Meanwhile, $\Ft_{i\pm 1/2}$ are correction
fluxes that make the method second-order accurate:
\begin{align}\label{correction-fluxes}
  \tilde{F}_{i\pm 1/2}=\frac{1}{2}\sum_p|s_{i\pm 1/2}^p|\left(1-\frac{\Delta t}{\Delta x}|s_{i\pm 1/2}^p|\right)\tilde\W_{i\pm 1/2}^p.
\end{align}
Here $\tilde{\W}_{i\pm 1/2}^p$ is a TVD-limited version of $\W_{i\pm 1/2}^p$.
We will sometimes work with
the first-order method obtained by setting the correction fluxes to zero:
\begin{align}\label{first-order_via_fluct}
  q_i^{n+1} = q_i^n-\frac{\Delta t}{\Delta x}\left[\apdq_\imh+\amdq_\iph\right]
\end{align}
which, for the conservative Riemann solvers we use herein, can also be written
as a first-order flux-differencing scheme:
\begin{align}\label{first-order_FV}
  q_i^{n+1}=q_i^n-\frac{\Delta t}{\Delta x}\left[\bfF_{i+1/2}-\bfF_{i-1/2}\right],
\end{align}
where $\bfF_{i+1/2}$ is a numerical approximation of the flux at $x_\iph$.

@@

In what follows, we use $\W^{p,\Rus}$ to denote the waves in the Rusanov
solver and simply $\W^p$ to denote the waves in the Roe solver.
%We close this section by remarking that the fluctuations associated with Rusanov's solver
%can be expressed in terms of waves and speeds from Roe's solver. To see this, first consider \eqref{rus_cons}
%and rewrite the right-going fluctuation as follows:
%\begin{align}\label{rus_rs_aux}
%  \A^{+,\Rus}\Delta Q_{i-1/2}:=\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\max}
%  = \bff(q_{i})-\bff(q_{i-1}) + \lambda_{i-1/2}^{\max}\left(\W_{i-1/2}^{1,\Rus}+\W_{i-1/2}^{2,\Rus}\right)
%  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus}.
%  %\A^{+,\Rus}\Delta Q_{i-1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p
%\end{align}
%From \eqref{system_for_alphas} and \eqref{rusanov_waves} we get
%$\W^{1,\Rus}_{i-1/2}+\W^{2,\Rus}_{i-1/2}=q_i-q_{i-1}=\sum_p\W_{i-1/2}^p$.
%Using \eqref{rs_conservation}, \eqref{rus_rs_aux} becomes
%\begin{align*}
%  \lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\max}
%  = \sum_p \left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p
%  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus},
%\end{align*}
%which implies that
%\begin{subequations}\label{fluct_rus}
%\begin{align}
%  \A^{+,\Rus}\Delta Q_{i-1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p,
%\end{align}
%and similarly,
%\begin{align}
%  \A^{-,\Rus}\Delta Q_{i+1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i+1/2}^p - \lambda_{i+1/2}^{\max}\right)\W_{i+1/2}^p.
%\end{align}
%\end{subequations}

@@

\subsection{The entropy residual}
Let $\entvar(\bfq)$ and $\entflux(\bfq)$ denote the entropy variable and entropy flux.
Based on \cite{guermond2011entropy}, we consider
\begin{align}\label{ent_residual}
  \int_{S_i} \entvar(\bfq) \cdot \left[ \frac{\partial \bfq}{\partial t} + \nabla\cdot \bff(\bfq)\right]d\bfx
  =\int_{S_i} \left[\frac{\partial \eta(\bfq)}{\partial t} + \entvar(\bfq) \cdot \nabla\cdot \bff(\bfq)\right] d\bfx
\end{align}
as a measurement of the entropy production in cell $i$ (here $S_i=[x_\imh,x_\iph]$.
To avoid the need to compute the time derivative of the entropy, we follow
\cite{guermond2018second, guermond2018well} and use
$\int \frac{\partial\eta(\bfq)}{\partial t}d\bfx=-\int \nabla\cdot\entflux(\bfq) d\bfx$,
which holds in smooth regions. Then we define
\begin{align}\label{Ri}
  \theta_i := \frac{R_i}{D_i},
\end{align}
with
\begin{align*}
  R_i=
  \left|\int_{S_i} \left[\entvar(\bfq) \cdot \nabla\cdot \bff(\bfq) - \nabla\cdot\entflux(\bfq) \right] d\bfx\right|,
\end{align*}
and $D_i$ being an upper bound of $R_i$ that acts as normalization so that $0\leq \theta_i\leq 1$.
Note that $R_i\approx 0$ if $\bfq$ is smooth in $S_i$.
In our implementation, we use
\begin{align*}
  R_i\approx \left|\entvar(\bfq_i)\cdot \int_{S_i}\nabla\cdot \bff(\bfq)d\bfx
  -\int_{S_i}\nabla\cdot\entflux(\bfq) d\bfx\right|,
  \quad
  %D_i = ||\entvar(\bfq_i)||_{\ell^2}\left|\left|\int_{S_i}\nabla\cdot\bff(\bfq)d\bfx\right|\right|_{\ell^2}
  %+\left|\int_{S_i}\nabla\cdot\bfq(\bfq)d\bfx\right|.
  D_i = \sum_{k=1}^{d+1}|\entvar_k(\bfq_i)|\left|\int_{S_i}\left(\nabla\cdot\bff(\bfq)\right)_kd\bfx\right|
  +\left|\int_{S_i}\nabla\cdot\entflux(\bfq)d\bfx\right|,
\end{align*}
where $(z)_k$ denotes the $k$-th component of $z\in\mathbb{R}^{d+1}$ and $d$ is the number of physical
dimensions. In the next section, we use $\theta_i$ to blend Rusanov's and Roe's Riemann solvers.

@@

To implement a second-order LWL scheme based on this solver, in addition to
these fluctuations we must define a set of waves and corresponding speeds.
Using only the waves from the Roe solver, it is in general not possible
to choose speeds that yield the fluctuations \eqref{ev_fluctuations}.
%In the framework of LeVeque, for a given interface $i+1/2$,
%we would need to consider a wave speed $s_{i+1/2}^p$ so that
%\begin{align*}
%  \hat \lambda_{i+1/2}^p+\lambda_{i+1/2}^p = s_{i+1/2}^p+|s_{i+1/2}^p|, \qquad
%  \hat \lambda_{i+1/2}^p-\lambda_{i+1/2}^p = s_{i+1/2}^p-|s_{i+1/2}^p|,
%\end{align*}
%which is not possible to obtain in general (since the problem is overdetermined).
Instead, we consider
$s_{i+1/2}^p=\sgn(\hat\lambda_{i+1/2}^p)\lambda_{i+1/2}^p$
 and study the consequences of this choice once the second-order corrections are applied.
Here $\sgn(z)$ is the sign function of $z$.

The second-order LWL method for the one-dimensional problem is given by
\eqref{second-order_via_fluct} and \eqref{correction-fluxes}.
To understand the effect of choosing $s_{i+1/2}^p=\sgn(\hat\lambda_{i+1/2}^p)\lambda_{i+1/2}^p$
as the speed of $\W_{i+1/2}^p$,
let us assume that no limiter is applied; i.e., $\tilde\W_{i\pm 1/2}^p\equiv \W_{i\pm 1/2}^p$.
Then \eqref{second-order_via_fluct} can be written as
\begin{align*}
  q_i^{n+1}=q_i^n
  &-\frac{\Delta t}{2 \Delta x}
  \sum_p
  \left[\left(\hat\lambda_{i+1/2}^p -\frac{\Delta t}{\Delta x}(\lambda_{i+1/2}^p)^2 \right)\W_{i+1/2}^p
  +
  \left(\hat\lambda_{i-1/2}^p +\frac{\Delta t}{\Delta x}(\lambda_{i-1/2}^p)^2 \right)\W_{i-1/2}^p\right],
\end{align*}
which is just the standard Lax-Wendroff method written wave propagation form.
Now taking $\theta_{i\pm 1/2}=0$ corresponds to using a Roe solver while taking
$\theta_{i\pm 1/2}=1$ corresponds to using a Rusanov solver.

Note that if $\theta_{i\pm 1/2}=0$ we recover the standard second-order Lax-Wendroff method based on
Roe's Riemann solver.
In the numerical experiments in \S\ref{sec:num}, we obtain second-order experimental
order of convergence (EOC) for problems with smooth solution, which is expected since
$\theta_{i\pm 1/2}\approx 0$ in smooth regions.
Similarly, if $\theta_{i\pm 1/2}=1$, we recover the Lax-Wendroff method based on Rusanov's Riemann solver,
which can be seen by writing the fluctuations in Rusanov's method in terms of Roe's waves and speeds.

To see this, first consider \eqref{rus_cons} and rewrite the right-going fluctuation as follows:
\begin{align}\label{rus_rs_aux}
  \A^{+,\Rus}\Delta Q_{i-1/2}:=\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus}
  = \bff(\bfq_{i})-\bff(\bfq_{i-1}) + \lambda_{i-1/2}^{\max}\left(\W_{i-1/2}^{1,\Rus}+\W_{i-1/2}^{2,\Rus}\right)
  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus}.
  %\A^{+,\Rus}\Delta Q_{i-1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p
\end{align}
From \eqref{system_for_alphas} and \eqref{rusanov_waves} we get
$\W^{1,\Rus}_{i-1/2}+\W^{2,\Rus}_{i-1/2}=\bfq_i-\bfq_{i-1}=\sum_p\W_{i-1/2}^p$.
Using \eqref{rs_conservation}, \eqref{rus_rs_aux} becomes
\begin{align*}
  \lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus}
  = \sum_p \left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p
  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus},
\end{align*}
which implies that
\begin{subequations}\label{fluct_rus}
\begin{align}
  \A^{+,\Rus}\Delta Q_{i-1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p,
\end{align}
and similarly,
\begin{align}
  \A^{-,\Rus}\Delta Q_{i+1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i+1/2}^p - \lambda_{i+1/2}^{\max}\right)\W_{i+1/2}^p.
\end{align}
\end{subequations}

@@

We consider a total of six problems. We start in \S \ref{sec:rp_dry_bed} with a
one-dimensional Riemann problem over a dry bed.
For this problem we use the first-order methods \eqref{first-order_via_fluct},
to avoid negative depth values. Although only Rusanov's solver is proven to guarantee
positivity, we do not get negative values for the water height with any of the first-order methods.
In \S \ref{sec:rp_wet_bed} we apply the second-order methods to a dam-break problem
with a wet bed.
In \S \ref{sec:steady_outflow} we consider a one-dimensional problem with a smooth steady state solution. We observe the expected
second-order accuracy of method \eqref{second-order_via_fluct} with Roe's and the blended solvers. Using
Rusanov's Riemann solver degrades the accuracy to first-order. The overall accuracy of the blended solver
is not degraded since the extra dissipation is not applied in smooth regions.
In \S\ref{sec:rp_wet_bed} and \S\ref{sec:1D_chj} we consider one-dimensional problems with strong shocks. We observe similar accuracy with the
blended solver or the Roe solver; this suggests that the numerical entropy
dissipation near the shocks does not degrade the accuracy
of the underlying Roe solver. This extra dissipation, however, prevents the formation of carbuncles in other experiments,
which we demonstrate in \S \ref{sec:bow_shock} and \S\ref{sec:2D_chj}.
The main focus of this work is in the formation of carbuncle instabilities in the two-dimensional CHJ. We present
an extensive set of experiments for this problem in \S \ref{sec:2D_chj}.
We consider not only Roe's, Rusanov's and the blended solvers, but also the HLLE solver and a solver proposed in \cite{kemm2014note},
which is designed to remove the carbuncle instabilities in the shallow water equations.

\subsection{Dam break problem on a dry bed}\label{sec:rp_dry_bed}
We start with the one-dimensional dam break problem on a dry bed.
This problem is a canonical example that demonstrates the `entropy glitch' of Roe's solver
at transonic rarefactions. Since the baseline Roe solver we use (see \S\ref{sec:roe}) does not contain
an entropy fix, it is important to demonstrate that the blended Riemann solver fixes the entropy glitch.
We follow the setup in \cite[\S 4.1.2]{delestre2013swashes}.
The domain is given by $\Omega=(0,L=10)$, and the initial condition is $hu(x,0)=0$ and
\begin{align*}
  h(x,0)=
  \begin{cases}
    h_l, & \mbox{ if }, x\leq x_0=5 \\
    h_r, &\mbox{ otherwise},
  \end{cases}
\end{align*}
with $h_l=5\times 10^{-3}$ and $h_r=0$. In our experiments we use $h_r=1\times 10^{-15}$ to avoid division by zero.
The exact solution, which can also be found in \cite{delestre2013swashes} and references therein, is
\begin{align*}
  h(x,t) =
  \begin{cases}
    h_l, \\
    \frac{4}{9g}\left(\sqrt{gh_l}-\frac{x-x_0}{2t}\right)^2, \\
    0,
  \end{cases}
\quad
  u(x,t) =
  \begin{cases}
    0, &\mbox{ if } x\leq x_A(t), \\
    \frac{2}{3}\left(\sqrt{gh_l}+\frac{x-x_0}{t}\right), & \mbox{ if } x_A(t) < x\leq x_B(t), \\
    0, &\mbox{ if } x_B(t)\leq x,
  \end{cases}
\end{align*}
with $x_A(t)=x_0-t\sqrt{gh_l}$ and $x_B(t)=x_0+2t\sqrt{gh_l}$.
We take $g=1$ and solve the problem up to $t=10$.
In Figures \ref{fig:rp_dry_bed_roe}-\ref{fig:rp_dry_bed_blended},
we show the solution using method \eqref{first-order_via_fluct}
with Roe's solver, Rusanov's solver and the blended solver, respectively.
The entropy glitch, which is manifested as a non-physical shock at $x=x_0$,
is present with Roe's solver. Using Rusanov's and the blended solvers
fixes the entropy glitch.
In Table \ref{table:rp_dry_bed}, we summarize the results of a convergence test.
Note that using the blended solver leads to more accurate results.

@@

For this particular problem, using method \eqref{first-order_via_fluct} with the blended Riemann solver
leads to $\lambda_{i\pm 1/2}^{\min}=0$ for all the experiments, but this is not true in general.
Finally, let us demonstrate the effect of $\lambda_{i\pm 1/2}^{\min}$ by imposing $\theta_{i}=0$ in \eqref{lambda_p},
which is equivalent to using Roe's solver with extra dissipation given by $\lambda_{i\pm 1/2}^{\min}$.
In Figure \ref{fig:rp_dry_bed_roe_with_lmin}, we show the solution using different refinements,
and in the last column of Table \ref{table:rp_dry_bed}, we summarize the results of a convergence test.
Note that the entropy glitch is greatly reduced.
To remove completely the entropy glitch we could add high-order entropy dissipation following \cite{tadmor2003entropy}
and references therein.
